Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-28 tex/draft.tex
edit llp section

\section{Long-lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs). \fxnote{incomplete sentence} That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at \fxnote{give ballparky number ($\approx$100um?)} to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from one of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay \fxnote{Mention that they have small couplings b/c weak force and high-mass intermediate particle is involved too}. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models meet different LLP criteria in different ways. The most common\fxnote{bad langauge. Andrew suggests 'most straightforward' or 'most ad-hoc'} SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from proton decay\fxnote{Be prepared to show this. Maybe add some diagrams illustrating proton decay w/o R-parity}. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as significant missing energy, making MET the conventional SUSY signature in many searches.

    On the other hand, models with R-parity violation (RPV) produce long-lived, but not perfectly stable LSPs when RPV terms have small coupling constants. Luckily, RPV models can still avoid proton decay by allowing either terms that violate baryon number or terms that violate lepton number, but not both. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs. \fxnote{is now a good time to explain GMSB?}\fxnote{cite dimopoulos low energy}

    LLPs from highly-virtual intermediate particles can occur in models such as split SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. \fxnote{cite kilian} \fxnote{mention that split throws out naturalness motivation?}\fxnote{Add diagrams}

    \fxnote{There is no zino. Maybe I meant bino? Neutralinos and charginos are the physical states that are linear combos of winos, bino, higgsinos} The degenerate mass spectra of other SUSY models can also give rise to LLPs. In some AMSB models \fxnote{explain AMSB?}, for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino then becomes long-lived when its decay is suppressed by the lack of available phase space. \fxnote{cite R\&S out of this world}
    \fxnote{merge last 2 paragraphs under 'mass spectra lead to LLP' concept?}
    \fxnote{never said anything about hidden valley or dark sector stuff. Never said the word R-hadron. Never said stealth or displaced SUSY. Should I?} 
    \fxnote{Andrew says don't worry about mentioning models that don't come up in the searches, but be prepared to explain the basic ideas behind HV, dark sector and how they result in LLP}

>>>>

\section{Long-lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs), which are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at \fxnote{give ballparky number ($\approx$100um?)} to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from any of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, on the other hand, are long-lived for reasons (2), (3), and (4). Both are pure weak decays, which gives them a smaller coupling constant, but they also decay through a W-boson, which is far more massive than either parent particle, and most importantly, the proton and electron masses leave them little phase space in which to decay. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}\fxnote{Add diagrams?}

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models produce LLPs in different ways. The most straightforward SUSY LLP occurs in models with conserved R-partity, a proposed symmetry that prevents proton decay\fxnote{Be prepared to show this. Maybe add some diagrams illustrating proton decay w/o R-parity}. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as missing energy (MET), making MET a key signature in many conventional searches.

    \fxnote{Diagrams showing some combination of proton decay, stable LSP, and RPV LSP}

    On the other hand, models with weakkly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs. Luckily, RPV models can still avoid proton decay by allowing either terms that violate baryon number or terms that violate lepton number, but not both\fxnote{cite something}. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs. \fxnote{is now a good time to explain GMSB?}\fxnote{cite dimopoulos low energy}

    LLPs can also arise from particular SUSY mass spectra. Models in the Split\fxnote{capitalize?}  SUSY paradigm, for example, propose that the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. \fxnote{cite kilian} \fxnote{mention that split throws out naturalness motivation?}\fxnote{Add diagrams} On the other hand, some AMSB models\fxnote{explain AMSB?} predict the zino\fxnote{no such thing. Maybe I meant bino or general neutralino?} and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino then becomes long-lived when its decay is suppressed by the lack of available phase space. \fxnote{cite R\&S out of this world}
    \fxnote{Andrew says don't worry about mentioning models that don't come up in the searches, but be prepared to explain the basic ideas behind HV, dark sector and how they result in LLP}
